---
title: "Creation of initial data sets for data analysis"
author: "Juraj Michalik"
date: "7/20/2022"
output: html_document
---

This will create initial data sets (separately for PBMCs from young and old mice) that has only base filtering.

Load needed packages.

Seurat: 4.0.3
R: 4.0.4
umap-learn (python, used by Seurat fci): 0.4.6

```{r}
library(Seurat)
library(data.table)
library(dplyr)
library(RColorBrewer)
library(grDevices)
library(stringr)
library(ggplot2)

source('VDJ_utilities.R')
```

Load data. After you perform mapping with Cell Ranger, put the path to *raw* matrix here.

```{r}
blood_young_tab <- Read10X('PBMC_B6_Young/raw_feature_bc_matrix/')
blood_young_VDJ <- load.10X.VDJ('PBMC_B6_Young_VDJ/all_contig_annotations.csv')

blood_old_tab <- Read10X('PBMC_B6_Old/raw_feature_bc_matrix/')
blood_old_VDJ <- load.10X.VDJ('PBMC_B6_Old_VDJ/all_contig_annotations.csv')
```

Load genes that should be present in final data. These genes were filtered on original combined data set, which also contained data from unrelated experiment (that was demultiplexed using cell surface markers and corrected cellular barcodes; see related paper and also GEO entry), using following filters:

* TR(AB)(VDJ) genes were filtered out;
* min.features = 200 (minimum number of unique genes in cell for the cell to be considered);
* **min.cells = 3** (gene must be present (at least 1 read) in at least 3 cells) - **this is the parameter which would make difference if applied to only extracted data**, as there are not all cells to which this parameter was applied).

Since we do not share also the other experiment data as they are part of unpublished project, we instead provide lists of genes that were present in final filtering when the data were still merged together, for the sake of reproducibility. You can also use min.cells = 3 filtering on this experiment only if you wish to - the results would be slightly different, but not by much - the conclusion would be same.

```{r}
genes_blood_young <- read.csv('genes/genes_BYB6.csv', header = F)[,1]
genes_blood_old <- read.csv('genes/genes_BOB6.csv', header = F)[,1]
```

Load filtered hashtags (generate these from script).

```{r}
hashtags_blood_young <- read.csv('hashtags/E02_Young_Blood_B6.csv', header = F)[,1]
hashtags_blood_old <- read.csv('hashtags/E02_Old_Blood_B6.csv', header = F)[,1]
```

# Blood Young Data

Prepare the Blood Young data (including antibodies)

```{r}
blood_young_tab$`Gene Expression` <- blood_young_tab$`Gene Expression`[rownames(blood_young_tab$`Gene Expression`) %in% genes_blood_young,]
B6by <- CreateSeuratObject(blood_young_tab$`Gene Expression`, project = 'B6 PBMC Young', min.features = 200)

antibodies <- blood_young_tab$`Antibody Capture`[,colnames(blood_young_tab$`Antibody Capture`) %in% colnames(B6by)]
antibodies <- CreateAssayObject(antibodies)
B6by[['Antibodies']] <- antibodies
```
Remove VDJ duplicates (either have more than 2CDR3 alpha sequences or more than 1 productive CDR3 beta sequence or more than 2 non-productive CDR3 beta sequences).

```{r}
vdj.doublets.by <- trace.VDJ.duplicates(B6by, blood_young_VDJ)
  
B6by <- subset(B6by, cells = colnames(B6by)[!(colnames(B6by) %in% vdj.doublets.by)])
B6by <- subset(B6by, cells = paste0(hashtags_blood_young, "-1"))
B6by$hashtags <- 'CD45-FB3'
B6by$age.group <- 'Young'

B6by <- create.VDJ.metadata(B6by, blood_young_VDJ)
B6by <- create.clones(B6by)
```

Write vdj data to Seurat object.

```{r}
B6by <- create.VDJ.metadata(B6by, blood_young_VDJ)
B6by <- create.clones(B6by)
```

Run initial analysis for Blood young data.
  
```{r}  
# run base processing
B6by <- NormalizeData(B6by, verbose = F) 
B6by <- FindVariableFeatures(B6by, verbose = F)
B6by <- ScaleData(B6by, verbose = F, features = rownames(B6by))
  
B6by <- RunPCA(B6by, features = VariableFeatures(B6by))
B6by <- RunUMAP(B6by, n.neighbors = 40, dims = 1:15)
  
B6by <- FindNeighbors(B6by, nn.method = 'rann', dims = 1:15)
B6by <- FindClusters(B6by, random.seed = 42)
```

```{r}
DimPlot(B6by)
```

# Blood Old Data

```{r}
blood_old_tab$`Gene Expression` <- blood_old_tab$`Gene Expression`[rownames(blood_old_tab$`Gene Expression`) %in% genes_blood_old,]
B6bo <- CreateSeuratObject(blood_old_tab$`Gene Expression`, project = 'B6 PBMC Old', min.features = 200)

antibodies <- blood_old_tab$`Antibody Capture`[,colnames(blood_old_tab$`Antibody Capture`) %in% colnames(B6bo)]
antibodies <- CreateAssayObject(antibodies)
B6bo[['Antibodies']] <- antibodies
```

Remove VDJ duplicates in the same way as before (reminder: either have more than 2CDR3 alpha sequences or more than 1 productive CDR3 beta sequence or more than 2 non-productive CDR3 beta sequences).

```{r}
vdj.doublets.bo <- trace.VDJ.duplicates(B6bo, blood_old_VDJ)
  
B6bo <- subset(B6bo, cells = colnames(B6bo)[!(colnames(B6bo) %in% vdj.doublets.bo)])
B6bo <- subset(B6bo, cells = paste0(hashtags_blood_old, "-1"))
B6bo$hashtags <- 'CD45-FB5'
B6bo$age.group <- 'Old'

B6bo <- create.VDJ.metadata(B6bo, blood_young_VDJ)
B6bo <- create.clones(B6bo)
```

Write vdj data to Seurat object.

```{r}
B6bo <- create.VDJ.metadata(B6bo, blood_old_VDJ)
B6bo <- create.clones(B6bo)
```

Run initial analysis for Blood young data.
  
```{r}  
# run base processing
B6bo <- NormalizeData(B6bo, verbose = F) 
B6bo <- FindVariableFeatures(B6bo, verbose = F)
B6bo <- ScaleData(B6bo, verbose = F, features = rownames(B6bo))
  
B6bo <- RunPCA(B6bo,features = VariableFeatures(B6bo))
B6bo <- RunUMAP(B6bo, n.neighbors = 40, dims = 1:15)
  
B6bo <- FindNeighbors(B6bo, nn.method = 'rann', dims = 1:15)
B6bo <- FindClusters(B6bo, random.seed = 42)
```

```{r}
DimPlot(B6bo)
```

Write out both data sets.

```{r}
dir.create('Datasets', showWarnings = F)
saveRDS(B6by, 'Datasets/Exp02_B6_Blood_Young_initial.rds')
saveRDS(B6bo, 'Datasets/Exp02_B6_Blood_Old_initial.rds')
```
